---
title: 'Homework: iterative method'
author: Athena Chen
output: html_notebook
---

```{r setup, include = FALSE}
source(file.path("..", "R", "util.R"))
required_packages <- c('RSpectra', 'testit')
install_and_load_packages(required_packages)
```

# Problem 1
Auto regressive processes can be viewed as a discrete analog of Ornstein–Uhlenbeck process &mdash; which coincides with Gaussian process based on an exponential covariance matrix &mdash; and hence is an example of Gaussian Markov random fields.
For instance, stationary lag-1 auto-regressive process 
$$x_t = \phi x_{t - 1} + \sqrt{1 - \phi^2} \, \epsilon_t, 
  \quad x_0 \sim \mathcal{N}(0, 1), 
  \quad \epsilon_t \mathbin{\overset{\small \textrm{i.i.d.}}{\sim}} \mathcal{N}(0, 1)$$
has the _tri-diagonal_ precision matrix
$$\boldsymbol{\Sigma}^{-1} = \frac{1}{1 - \phi^2} 
  \begin{bmatrix} 
  1 & -\phi & 0 & & & \ldots & 0 \\
  -\phi & 1 + \phi^2 & -\phi & 0 & & & \vdots \\
  0 & -\phi & 1 + \phi^2 & -\phi & 0 & & \\
    & & \ddots & \ddots & \ddots & & \\
    & & 0 & -\phi & 1 + \phi^2 & -\phi & 0 \\
  \vdots & &   & 0 & -\phi & 1 + \phi^2 & -\phi \\
  0 & \ldots &   &   & 0 & -\phi & 1\\
  \end{bmatrix}.$$
More generally, a lag-$k$ (non-stationary) auto-regressive process has a _banded_ precision matrix with bandwidth $k$.

Implement a fast matrix-vector $\boldsymbol{v} \to \boldsymbol{\Sigma}^{-1} \boldsymbol{v}$ operation, exploiting the structure of the AR-1 precision matrix.
Then use this function to find the top 10 principal components of $\boldsymbol{\Sigma}$ (not $\boldsymbol{\Sigma}^{-1}$) via Lanczos algorithm provided via `RSpectra::eigs_sym`.

```{r}
ar_length <- 4096
auto_corr <- .9 # Corresponds to `\phi` above

ar_precision_matvec <- function(v, auto_corr) {
  # Fill in: note that you can vectorize the calculation and do *not* need a for-loop. 
  # (Hint: how would you efficiently carry out a matrix-vector operation if the matrix has non-zero entries only along the sub or super diagonal?)
  n <- length(v)
  
  # pre-allocate vector
  v_out <- numeric(n)
  
  # Calculate first term
  v_out[1] <- v[1] - auto_corr * v[2]
  
  # Calculate middle terms
  v_mat <- cbind(v[1:(n-2)], v[2:(n-1)], v[3:n])
  v_out[2:(n-1)] <- v_mat %*% c(-auto_corr, 1 + auto_corr^2, -auto_corr)
  
  # Calculate last term
  v_out[n] <- -auto_corr * v[(n-1)] + v[n]
  
  # Adjust by constant
  v_out/(1 - auto_corr^2)
}

ar_eig <- eigs_sym(
  ar_precision_matvec, args = auto_corr,
  # Fill in
  k = 10, 
  which = "SM",
  opts = list(
    ncv = 100, # Spectrum distribution of AR-1 process is not very spread out on the extreme ends and is actually a hard case for Lanczos. So it helps to have more Lanczos vectors than the default for faster convergence.
    maxitr = 10^3, # Cap it just in case
    retvec = TRUE # More efficient to do without eigenvectors when not needed
  ),
  n = ar_length
)
```

Now, directly compute the eigen decomposition of $\boldsymbol{\Sigma}$ (not $\boldsymbol{\Sigma}^{-1}$) and compare its output with the principal components and associated variances found via Lancsoz algorithm.

<font color="blue"> It can be shown that for a stationary lag-1 AR process, 

$$ Cov(X_t, X_s) = \phi^{|s - t|} $$

Note that `ar_eig` contains the 10 smallest eigenvalues of $\Sigma^{-1}$ which are equal to the inverse of the ten largest eigenvalues of $\Sigma$. To show that the associated variances between the two algorithms are similar, it is sufficient to show that the eigenvalues are similar. Additionally, the eigen vectors may differ by the sign, so in the comparison we only look at the absolute value of the eigen vectors. In some cases, the eigen vectors derived the Lancos algorithm were fairly different from those derived directly. 
</font>

```{r}
Sigma <- sapply(1:ar_length, function(x) auto_corr^abs(x - 1:ar_length))
Sigma_ed <- eigen(Sigma)

lancos_eigen_val <- sort(1/ar_eig$values, decreasing = TRUE)
lancos_eigen_vec <- ar_eig$vectors[, order(1/ar_eig$values, decreasing = TRUE)]

# Check that that the two are similar
assert(paste0("The eigen decomposition obtained via the Lancsoz algorithm, ", 
              "is consistent with the result obtained directly."), {
                are_all_close(lancos_eigen_val, Sigma_ed$values[1:10])
                are_all_close(abs(lancos_eigen_vec), abs(Sigma_ed$vectors[, 1:10]), 
                              rel_tol = 5e-2)
})
```

**Remark:** 
For banded matrices, there actually are even more efficient approaches.
To get a sense of special routines available for banded matrices, you can take a look at `*_banded` functions in [SciPy's linear algebra routines](https://docs.scipy.org/doc/scipy/reference/linalg.html).
Even those functions represent only a subset of available numerical linear algebra techniques; see
[LAPACK documentation for SVD](https://www.netlib.org/lapack/lug/node32.html) for example.
